---
title: "REPRORODUCIBLE RESEARCH WITH BITCOIN FORECASTING"
author: "AYOUB JARBOUAI (418035) &&  IDRISS TALBI (418232)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
  word_document:
    toc: true
  pdf_document:
    toc: true
editor_options:
  markdown:
    wrap: sentence
---

# BITCOIN

------------------------------------------------------------------------


Bitcoin, introduced in 2009 by the mysterious Satoshi Nakamoto, stands as the pioneering cryptocurrency.
It operates on a decentralized peer-to-peer network, allowing secure and transparent transactions without the need for a central authority.
Bitcoin's blockchain technology records every transaction in a public ledger, ensuring immutability and trustworthiness through cryptographic hashing.

Our project is structured into two key sections:

**1. Data Exploration** - Analyzing and visualizing Bitcoin price data.
- Identifying trends and patterns in historical price movements.

**2. Time Series Analysis and Forecasting** - **Forecasting Methods:** - **LSTM (Long Short-Term Memory):** Leveraging deep learning to capture complex temporal dependencies.
- **XGBoost:** Harnessing gradient boosting to predict Bitcoin price fluctuations.
- **Facebook Prophet:** Using a versatile tool for forecasting seasonality and holidays.
- **ARIMA (AutoRegressive Integrated Moving Average):** Employing a traditional statistical approach for time series forecasting.

This project draws inspiration from notable kernels and tutorials, including those focused on Prophet, XGBoost, and ARIMA models, tailored specifically for Bitcoin price prediction.

For those new to Bitcoin, you can learn more about its fundamentals [here](https://www.coindesk.com/information/what-is-bitcoin).



------------------------------------------------------------------------





# Checking out the dataset in my data folder

From the code below  we set a default data folder we run the code and list the data files in the folder. For our case, it's

-   "bitstampUSD_1-min_data_2012-01-01_to_2018-11-11.csv"

This file has the daily data from 2012 first of January to 2018 eleventh of November hourly market prices for BTC

-   "coinbaseUSD_1-min_data_2014-12-01_to_2018-11-11.csv"

This file has the daily data from 2014 first of January to 2018 eleventh of November hourly market prices for BTC

```{r, message=FALSE}
# Set the directory path where the data is located
data_dir <- "~/Git/Data"

# List files in the directory specified by data_dir
dir_contents <- list.files(data_dir)

# Print the contents of the directory to the console
print(dir_contents)

```

# Environment preparation

## Getting packages

```{r, message=FALSE}
# Define the list of package names
packages <- c("tidyverse", "plotly", "lubridate", "readr", "dplyr", "zoo", "ggplot2", "forecast", "stats", "tseries")
# Use lapply to load each package
lapply(packages, require, character.only = TRUE)
```
