---
title: "REPRORODUCIBLE RESEARCH WITH BITCOIN FORECASTING"
author: "AYOUB JARBOUAI (418035) &&  IDRISS TALBI (418232)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
  word_document:
    toc: true
  pdf_document:
    toc: true
editor_options:
  markdown:
    wrap: sentence
---

# BITCOIN

------------------------------------------------------------------------


Bitcoin, introduced in 2009 by the mysterious Satoshi Nakamoto, stands as the pioneering cryptocurrency.
It operates on a decentralized peer-to-peer network, allowing secure and transparent transactions without the need for a central authority.
Bitcoin's blockchain technology records every transaction in a public ledger, ensuring immutability and trustworthiness through cryptographic hashing.

Our project is structured into two key sections:

**1. Data Exploration** - Analyzing and visualizing Bitcoin price data.
- Identifying trends and patterns in historical price movements.

**2. Time Series Analysis and Forecasting** - **Forecasting Methods:** - **LSTM (Long Short-Term Memory):** Leveraging deep learning to capture complex temporal dependencies.
- **XGBoost:** Harnessing gradient boosting to predict Bitcoin price fluctuations.
- **Facebook Prophet:** Using a versatile tool for forecasting seasonality and holidays.
- **ARIMA (AutoRegressive Integrated Moving Average):** Employing a traditional statistical approach for time series forecasting.

This project draws inspiration from notable kernels and tutorials, including those focused on Prophet, XGBoost, and ARIMA models, tailored specifically for Bitcoin price prediction.

For those new to Bitcoin, you can learn more about its fundamentals [here](https://www.coindesk.com/information/what-is-bitcoin).



------------------------------------------------------------------------





# Checking out the dataset in my data folder

From the code below  we set a default data folder we run the code and list the data files in the folder. For our case, it's

-   "bitstampUSD_1-min_data_2012-01-01_to_2018-11-11.csv"

This file has the daily data from 2012 first of January to 2018 eleventh of November hourly market prices for BTC

-   "coinbaseUSD_1-min_data_2014-12-01_to_2018-11-11.csv"

This file has the daily data from 2014 first of January to 2018 eleventh of November hourly market prices for BTC

```{r, message=FALSE}
# Set the directory path where the data is located
data_dir <- "~/Git/Data"

# List files in the directory specified by data_dir
dir_contents <- list.files(data_dir)

# Print the contents of the directory to the console
print(dir_contents)

```

# Environment preparation

## Getting packages

```{r, message=FALSE}
# Define the list of package names
packages <- c("tidyverse", "plotly", "lubridate", "readr", "dplyr", "zoo", "ggplot2", "forecast", "stats", "tseries")
# Use lapply to load each package
lapply(packages, require, character.only = TRUE)
```



## Data importation

-   **The data importation utilizes the read_csv function from the readr library. our Data has a Timestamp which will be converted to readable form with the code line `mutate(Timestamp = as.POSIXct(Timestamp, origin = "1970-01-01", tz = "UTC"))`**

-   The line interprets the Timestamp in a human-readable form.
    We set the format to UTC and then give it a sample date to use like in our case 1970-01-01.
    and set the path to our file.

-   Here we end the code by displaying the first 6 rows of data.

```{r, message=FALSE}
# Define a function to parse timestamps from seconds since the Unix epoch
dateparse <- function(time_in_secs) {
  as_datetime(as.numeric(time_in_secs), origin = "1970-01-01", tz = "UTC")
}

# Expand the tilde to the home directory and specify the file path
file_path <- "~/Git/Data/coinbaseUSD_1-min_data_2014-12-01_to_2018-11-11.csv"

# Read the CSV file into a data frame named 'data'
data <- read_csv("~/Git/Data/coinbaseUSD_1-min_data_2014-12-01_to_2018-11-11.csv")

# Convert the 'Timestamp' column from numeric Unix timestamps to POSIXct date-time objects
data <- data %>%
  mutate(Timestamp = as.POSIXct(Timestamp, origin = "1970-01-01", tz = "UTC"))

# Display the first few rows of the 'data' data frame
#head(data)

```



### Summary Statistics

Summary statistics display the minimum and maximum values, explaining the range.
They also include the 1st quantile (Q1), median (Q2), and third quantile (Q3), which elaborate on the interquartile range.
The summary also highlights the total number of missing variables in each column.

```{r}
summary(data)
```



## Data Cleaning

### Data handling Comparison with python



| **`R Code`**                                                                      | **`Python Code`**                                                                      |
|-------------------------------------|-----------------------------------|
| `r`                                                                               | `python`                                                                               |
| \# Load required libraries                                                        | \# Import necessary libraries                                                          |
| library(dplyr)                                                                    | import pandas as pd                                                                    |
| library(zoo)                                                                      | import numpy as np                                                                     |
|                                                                                   |                                                                                        |
| \# Sample data                                                                    | \# Sample data                                                                         |
| data \<- data.frame(                                                              | data = pd.DataFrame({                                                                  |
| Timestamp = seq.POSIXt(from = as.POSIXct("2022-01-01 00:00"),                     | 'Timestamp': pd.date_range(start='2022-01-01', periods=10, freq='H'),                  |
| by = "hour", length.out = 10),                                                    | 'Volume\_(BTC)': [0.5, np.nan, 0.6, np.nan, np.nan, 0.7, 0.8, np.nan, 0.9, 1.0],       |
| `Volume_(BTC)` = c(0.5, NA, 0.6, NA, NA, 0.7, 0.8, NA, 0.9, 1.0),                 | 'Volume\_(Currency)': [500, np.nan, 600, np.nan, np.nan, 700, 800, np.nan, 900, 1000], |
| `Volume_(Currency)` = c(500, NA, 600, NA, NA, 700, 800, NA, 900, 1000),           | 'Weighted_Price': [100, np.nan, 102, np.nan, np.nan, 105, 106, np.nan, 108, 109],      |
| Weighted_Price = c(100, NA, 102, NA, NA, 105, 106, NA, 108, 109),                 | 'Open': [100, 101, np.nan, 103, np.nan, 105, 106, np.nan, 108, 109],                   |
| Open = c(100, 101, NA, 103, NA, 105, 106, NA, 108, 109),                          | 'High': [101, 102, np.nan, 104, np.nan, 106, 107, np.nan, 109, 110],                   |
| High = c(101, 102, NA, 104, NA, 106, 107, NA, 109, 110),                          | 'Low': [99, 100, np.nan, 102, np.nan, 104, 105, np.nan, 107, 108],                     |
| Low = c(99, 100, NA, 102, NA, 104, 105, NA, 107, 108),                            | 'Close': [100, 101, np.nan, 103, np.nan, 105, 106, np.nan, 108, 109]                   |
| Close = c(100, 101, NA, 103, NA, 105, 106, NA, 108, 109)                          | })                                                                                     |
| )                                                                                 |                                                                                        |
|                                                                                   |                                                                                        |
| \# Replace NA with 0 for volume/trade-related fields                              | \# Replace NA with 0 for volume/trade-related fields                                   |
| data \<- data %\>%                                                                | data['Volume\_(BTC)'].fillna(value=0, inplace=True)                                    |
| mutate(`Volume_(BTC)` = ifelse(is.na(`Volume_(BTC)`), 0, `Volume_(BTC)`),         | data['Volume\_(Currency)'].fillna(value=0, inplace=True)                               |
| `Volume_(Currency)` = ifelse(is.na(`Volume_(Currency)`), 0, `Volume_(Currency)`), | data['Weighted_Price'].fillna(value=0, inplace=True)                                   |
| Weighted_Price = ifelse(is.na(Weighted_Price), 0, Weighted_Price))                |                                                                                        |
|                                                                                   |                                                                                        |
| \# Fill forward NA values for OHLC data                                           | \# Fill forward NA values for OHLC data                                                |
| data$Open <- zoo::na.locf(data$Open)                                              | data['Open'].fillna(method='ffill', inplace=True)                                      |
| data$High <- zoo::na.locf(data$High)                                              | data['High'].fillna(method='ffill', inplace=True)                                      |
| data$Low <- zoo::na.locf(data$Low)                                                | data['Low'].fillna(method='ffill', inplace=True)                                       |
| data$Close <- zoo::na.locf(data$Close)                                            | data['Close'].fillna(method='ffill', inplace=True)                                     |
|                                                                                   |                                                                                        |
| \# Convert 'Timestamp' column to POSIXct                                          | \# Convert 'Timestamp' column to datetime                                              |
| data \<- data %\>%                                                                | data['Timestamp'] = pd.to_datetime(data['Timestamp'], unit='s', utc=True)              |
| mutate(Timestamp = as.POSIXct(Timestamp, origin = "1970-01-01", tz = "UTC"))      |                                                                                        |
|                                                                                   |                                                                                        |
| \# Display the first few rows of the updated dataframe                            | \# Display the first few rows of the updated dataframe                                 |
| head(data)                                                                        | print(data.head())                                                                     |
| `|`                                                                               |                                                                                        |

### Comments on Differences:

1.  **Library Usage**:
    -   R uses `dplyr` for data manipulation and `zoo` for forward filling.
    -   Python uses `pandas` for both data manipulation and forward filling.
2.  **NA Replacement**:
    -   In R, `ifelse` within `mutate` is used to replace NA values.
    -   In Python, `fillna` is used directly on the DataFrame columns.
3.  **Forward Filling**:
    -   R uses `zoo::na.locf` to forward fill values.
    -   Python uses `fillna(method='ffill')` for forward filling.
4.  **Timestamp Conversion**:
    -   R uses `as.POSIXct` to convert numeric Unix timestamps.
    -   Python uses `pd.to_datetime` with `unit='s'` and `utc=True` to achieve the same.

### Handling missing values

-   From the code below we fill in missing values in the Open High Low and Close columns and replace NAs with the previous value from the one above it.

-   This approach, in the beginning, helps to maintain the continuity of data without losing important rows of the data in other columns.

```{r}
# Fill NA values with zeroes for volume/trade-related fields using dplyr's mutate function
data <- data %>%
  mutate(`Volume_(BTC)` = ifelse(is.na(`Volume_(BTC)`), 0, `Volume_(BTC)`),  # Replace NA with 0 in 'Volume_(BTC)'
         `Volume_(Currency)` = ifelse(is.na(`Volume_(Currency)`), 0, `Volume_(Currency)`),  # Replace NA with 0 in 'Volume_(Currency)'
         Weighted_Price = ifelse(is.na(Weighted_Price), 0, Weighted_Price))  # Replace NA with 0 in 'Weighted_Price'

# Fill forward NA values (last observation carried forward) for OHLC data using zoo's na.locf function
data$Open <- zoo::na.locf(data$Open)    # Fill NA values in 'Open' column
data$High <- zoo::na.locf(data$High)    # Fill NA values in 'High' column
data$Low <- zoo::na.locf(data$Low)      # Fill NA values in 'Low' column
data$Close <- zoo::na.locf(data$Close)  # Fill NA values in 'Close' column

# Convert 'Timestamp' column from numeric Unix timestamps to POSIXct date-time objects in UTC timezone
data <- data %>%
  mutate(Timestamp = as.POSIXct(Timestamp, origin = "1970-01-01", tz = "UTC"))

# Display the first few rows of the updated dataframe 'data'
#head(data)
```

